---
title: "Nuclear membrane quantification of smFISH signal in U2OS Cells"
author: "Sam Zavislan-Pullaro and Erin Osborne Nishimura"
date: "2026-08-13"
output: 
  github_document:
    toc: true
    toc_depth: 4
---

# Step 1: libraries, import, and data munging

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Load libraries

  - tidyverse
  - data.table
  - rstatix
  - ggrepel
  - gridExtra

```{r setup}
library(tidyverse)
library(data.table)
library(rstatix)
library(ggrepel)
library(gridExtra)
```

---

## Import data

Note:     

"ch1" = gapdh — dark gray
"ch2" = tnpo-1 — dark blue

Strain: U2OS

```{r import}

# Load the data
# Only need to replace name of file. EX: ("../01_input/<NAME_OF_DATAFILE.txt>",header = FALSE, sep = "\t")

df_1 <- read.table("../01_input/2026-8-3_1411_datafile_for_231206_imaging_U2OS.txt",
                        header = FALSE, sep = "\t")
df_2 <- read.table("../01_input/2026-8-3_1419_datafile_for_250802A_TNPO1_U2OS_rep2.txt",
                        header = FALSE, sep = "\t")
df_3 <- read.table("../01_input/2026-8-3_1431_datafile_for_250802B_TNPO1_U2OS_rep3.txt",
                        header = FALSE, sep = "\t")

# Check dimensions
# In this case, all data files should have vector length = 183 This is determined by the length of the ROI box in FIJI

# dim(df_1)
# dim(df_2)
# dim(df_3)

df_1 <- df_1[,1:183]
df_2 <- df_2[,1:183]
df_3 <- df_3[,1:183]

# The value 181 is actually determined by the dimensions, just subtract 2 from what you get in dimensions

col_times <- paste(rep(1:181), sep = "") 

# Set column names for the dataframes 

colnames(df_1) <- c("file", "channel", col_times)
colnames(df_2) <- c("file", "channel", col_times)
colnames(df_3) <- c("file", "channel", col_times)
# dim(df_1)
```

---

## Create annotation columns and add timepoints

Timepoints are treated as positions along line scan

```{r annotate}

# Helper function: pivot longer and annotate 
annotate_df <- function(df) {
  df[2:nrow(df), 1:183] %>% #  comes from dimensions
    separate_wider_delim(file, delim = "_",
                         names = c("date", "strain", "mRNA1", "mRNA2", "embryoID", NA)) %>%
    pivot_longer(cols = `1`:`181`, names_to = "xpoint", values_to = "intensity") %>%  #change the dimension value according to line 63
    mutate(unique_id = paste(date, strain, mRNA1, mRNA2, embryoID, channel, sep = "_"))
}


# Helper function: add timepoints from the header row of any loaded table
addInTimepoints <- function(tibble, ref_table) {
  timepoints <- as.vector(unlist(ref_table[1, 3:183])) #need the full original dimension
  repnumb    <- nrow(tibble) / 181
  tibble::add_column(tibble, timepoints = rep(timepoints, repnumb), .after = "xpoint")
}

# Add timepoints as proxy for pixel position along line scan 
df_1_pt <- addInTimepoints(annotate_df(df_1), df_1) # Use first dataframe as reference 
df_2_pt <- addInTimepoints(annotate_df(df_2), df_1)
df_3_pt <- addInTimepoints(annotate_df(df_3), df_1)

# Sanity check 
#head(df_1_pt)
#df_1_pt[180:190,]

```

---

## Merge all data and apply fixed coordinate alignment


```{r merge_align}

# Concatenate the dataframes 
all_data <- rbind(
  df_1_pt, df_2_pt, df_3_pt)

center <- ceiling(181/2) # Using dims, find center point 
center

dt <- as.data.table(all_data)
dt$xpoint <- as.integer(dt$xpoint)
dt[, aligned_row := xpoint - center]

total_align_long <- dt %>%
  select(strain, aligned_row, unique_id, intensity)

cat("Total rows:", nrow(total_align_long), "\n")
cat("Unique embryo IDs:", n_distinct(total_align_long$unique_id), "\n")

# head(total_align_long)

```

---

# Step 2: Normalize

Each embryo is normalized to its own mean intensity (per embryo per channel).

```{r normalize}

norm_total <- total_align_long %>%
  separate_wider_delim(unique_id, delim = "_",
                       names = c("date", NA, "mRNA1", "mRNA2", "embryoID", "channel")) %>%
  group_by(date, embryoID, channel, mRNA1, mRNA2) %>%
  mutate(normalized_intensity = intensity / mean(intensity, na.rm = TRUE)) %>%
  ungroup()

treatment_order <- c("ch1", "ch2")
norm_total$channel <- factor(norm_total$channel, levels = treatment_order)

# ch1 = gapdh, ch2 = tnpo-1
norm_total <- norm_total %>%
  mutate(treatment_type = case_when(
    channel == "ch1" ~ "gapdh",
    channel == "ch2" ~ "tnpo-1",
    TRUE ~ NA_character_
  ))

table(norm_total$channel, norm_total$treatment_type)

#head(norm_total)

```

---

## Color palettes

```{r palettes}

treatment_colors <- c(
  "ch2" = "#2166AC",   # tnpo-1 — dark blue
  "ch1" = "#4D4D4D"    # gapdh — dark gray
)
type_colors <- treatment_colors

channel_labels <- c("ch1" = "gapdh mRNA", "ch2" = "tnpo-1 mRNA")

```

---

# Step 3: Generate individual normalized line scans

All embryo-level line scans, faceted by channel and treatment.

```{r linescans_individual, fig.height = 8, fig.width = 10}

p_linescan_individual <- ggplot(norm_total,
       aes(x = aligned_row, y = normalized_intensity,
           group = interaction(date, embryoID),
           color = channel)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_manual(values = treatment_colors) +
  facet_wrap(~ channel, labeller = labeller(channel = channel_labels)) +
  labs(
    title = "Individual normalized line scans — all treatments",
    x     = "Position along line scan",
    y     = "Normalized intensity",
    color = "Treatment"
  ) +
  guides(color = "none") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

p_linescan_individual


```

---

## Save the individual linescan plots

```{r plotsave1}

# Create an output directory
output_dir <- "../03_output"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Save a filename
today <- format(Sys.Date(), "%y%m%d")
filename <- paste(output_dir, "/", today, "_linescan_individual_U2OS.pdf", sep = "")

# Save the plot
pdf(filename, width = 7, height = 6)
p_linescan_individual
dev.off()

```

---

# Step 4: Generate mean line scans

```{r linescans_mean_overlaid, fig.height = 6, fig.width = 12}

mean_signal <- norm_total %>%
  group_by(aligned_row, channel) %>%
  summarize(
    mean_signal = mean(normalized_intensity, na.rm = TRUE),
    sd_signal   = sd(normalized_intensity,   na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  mutate(
    ymeanmax = mean_signal + sd_signal,
    ymeanmin = mean_signal - sd_signal
  )

# Plot the means and sd as linescans and ribbons
p_linescan_mean <- ggplot(mean_signal,
       aes(x = aligned_row, y = mean_signal,
           group = channel)) +
  geom_ribbon(aes(ymin = ymeanmin, ymax = ymeanmax,
                  fill = channel), alpha = 0.2, color = NA) +
  geom_line(aes(color = channel), linewidth = 0.8) +
  scale_color_manual(values = treatment_colors, labels = channel_labels) +
  scale_fill_manual(values  = treatment_colors, labels = channel_labels) +
  labs(
    title = "Mean +- SD line scans, per channel",
    x     = "Position",
    y     = "Mean normalized intensity"
  ) +
  theme_bw(base_size = 11)

p_linescan_mean

```

---

## Save the mean line plots


```{r plotsave2}

# Save a filename
today <- format(Sys.Date(), "%y%m%d")
filename <- paste(output_dir, "/", today, "_linescan_mean_U2OS.pdf", sep = "")

# Save the plot
pdf(filename, width = 7, height = 6)
p_linescan_mean
dev.off()

```
---

# Step 5: Calculate log2 fold change enrichment

Ratio of normalized intensity at position 1 versus the mean of positions
-50 and +50. Higher values indicate more enriched localization signal
(consistent with membrane localization).

```{r}

peaks_and_valleys <- norm_total %>%
  filter(aligned_row %in% c(-50, 1, 50))

nest_pandv <- peaks_and_valleys %>%
  group_by(date, mRNA1, mRNA2, embryoID, channel) %>%
  nest()

my_calc2 <- function(df) {
  df$normalized_intensity[df$aligned_row == 1] /
    mean(df$normalized_intensity[df$aligned_row %in% c(-50, 50)])
}

foldChange_calc <- nest_pandv %>%
  mutate(fc_enrich = map_dbl(data, my_calc2)) %>%
  mutate(
    log2_fc        = log(fc_enrich, base = 2),
    treatment_type = case_when(
      channel == "ch1" ~ "gapdh",
      channel == "ch2" ~ "tnpo-1",
      TRUE ~ NA_character_
    ),
    channel        = factor(channel, levels = treatment_order)
  )

summary(foldChange_calc$log2_fc)

```

## Generate merged boxplot, x-axis by treatment

```{r }
foldChange_calc <- foldChange_calc %>%
  mutate(treatment_type = factor(treatment_type, levels = c("gapdh", "tnpo-1")))

```


---

## Summary metrics

```{r summary_stats}

summary_stats_table <- foldChange_calc %>%
  group_by(treatment_type, channel) %>%
  summarise(
    n             = n(),
    min_log2fc    = round(min(log2_fc,    na.rm = TRUE), 3),
    median_log2fc = round(median(log2_fc, na.rm = TRUE), 3),
    mean_log2fc   = round(mean(log2_fc,   na.rm = TRUE), 3),
    max_log2fc    = round(max(log2_fc,    na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(channel, treatment_type)

print(summary_stats_table, n = Inf)

```
---

## Statistics - Pairwise Wilcoxon

```{r pairwise_wilcox}

# Wilcoxon test comparing set-3 vs imb-2 enrichment
pairwise_stats <- foldChange_calc %>%
  ungroup() %>%
  wilcox_test(log2_fc ~ treatment_type) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

pairwise_stats

# pairwise_stats$p

```
---

## Reps and N-values

How many replicates and samples are represented in the dataset?

```{r reps}
      
# select the metadata
metadata <- norm_total %>% 
  select("strain", "date", "embryoID")

# pick only the unique ones
metadata <- unique(metadata)

# dim(metadata)
# str(metadata)
# head(metadata)

# tabulate the number of n-values.
rep_and_n <- table(metadata$date)
rep_and_n

```

## Plot boxplot

```{r boxplot_merged, fig.height = 6, fig.width = 8}

p_merged <- ggplot(foldChange_calc,
                   aes(x = treatment_type,
                       y = log2_fc,
                       fill = treatment_type)) +
  geom_boxplot(outlier.shape = NA, width = 0.6) +
  geom_jitter(aes(color = treatment_type),
              position = position_jitterdodge(jitter.width = 0.25),
              size = 1.5, alpha = 0.7) +
  scale_fill_manual(
    values = c("tnpo-1" = "#2166AC", "gapdh" = "#4D4D4D"),
    guide  = "none"
  ) +
  scale_color_manual(
    values = c("tnpo-1" = "black", "gapdh" = "black"),
    guide  = "none"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_y_continuous(limits = c(-1, 1)) +
  labs(
    title = "log2 enrichment at position 0 vs ±50",
    x     = "mRNA",
    y     = "log2 (intensity at 0 / mean intensity at ±50)",
    fill  = NULL
  ) +
  theme_classic(base_size = 12) +
  annotate("text", x = 1.5, y = 1, label = paste("p = ", pairwise_stats$p, sep = "")) +
  theme(
    axis.text.x     = element_text(angle = 35, hjust = 1, face = "italic"),
    strip.text      = element_text(face = "bold"),
    legend.position = "none"
  )

p_merged

```

---

## Save the box plots


```{r plotsave3}

# Save a filename
today <- format(Sys.Date(), "%y%m%d")
filename <- paste(output_dir, "/", today, "_boxplot_merged_U2OS.pdf", sep = "")

# Save the plot
pdf(filename, width = 7, height = 6)
p_merged
dev.off()

```

# Step 6: Export plots and stats

```{r export_outputs}

# Save some files as .csvs
write_csv(dt, file.path(output_dir,paste0(today, "_aligned_raw_data_U2OS.csv")))
write_csv(summary_stats_table, file.path(output_dir, paste0(today, "_summary_stats_U2OS.csv")))
write_csv(pairwise_stats, file.path(output_dir, paste0(today, "_pairwise_wilcoxon_stats_U2OS.csv")))


# Save the rep and n nubmers
today <- format(Sys.Date(), "%y%m%d")
filename <- paste(output_dir, "/", today, "_rep_and_n.txt", sep = "")
write.table(rep_and_n, file = filename, sep = "\t", quote = FALSE, row.names = FALSE)


```

---

# Session info

```{r sessionInfo}
sessionInfo()
```

---