---
title: "Ambika: tnpo-1 gapdh nuclear membrane signal quantification and analysis"
author: "Sam Zavislan-Pullaro"
date: "2026-07-30"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: flatly
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Load libraries

```{r setup}
library(tidyverse)
library(data.table)
library(rstatix)
library(ggrepel)
library(gridExtra)
```

---

## Import data

Note: channel 1 = imb-2 mRNA | channel 2 = set-3 mRNA

Strain: N2 

```{r import}

# Load the data
# Only need to replace name of file. EX: ("../01_input/<NAME_OF_DATAFILE.txt>",header = FALSE, sep = "\t")

df_1 <- read.table("/Users/ambikabasu/Nishimura_lab/Macros/04_R_analysis_tnpo/01_input/2026-8-3_1411_datafile_for_231206_imaging_U2OS.txt",
                        header = FALSE, sep = "\t")
df_2 <- read.table("/Users/ambikabasu/Nishimura_lab/Macros/04_R_analysis_tnpo/01_input/2026-8-3_1419_datafile_for_250802_TNPO1_U2OS_rep2.txt",
                        header = FALSE, sep = "\t")
df_3 <- read.table("/Users/ambikabasu/Nishimura_lab/Macros/04_R_analysis_tnpo/01_input/2026-8-3_1428_datafile_for_250802_TNPO1_U2OS_rep2.txt",
                        header = FALSE, sep = "\t")
df_4 <- read.table("/Users/ambikabasu/Nishimura_lab/Macros/04_R_analysis_tnpo/01_input/2026-8-3_1431_datafile_for_250802_TNPO1_U2OS_rep3.txt",
                       header = FALSE, sep = "\t")
df_5 <- read.table("/Users/ambikabasu/Nishimura_lab/Macros/04_R_analysis_tnpo/01_input/2026-8-3_1438_datafile_for_250802_TNPO1_U2OS_rep3.txt",
                        header = FALSE, sep = "\t")

# Check dimensions
# In this case, all data files should have vector length = 121. This is determined by the length of the ROI box in FIJI

dim(df_1)
dim(df_2)
dim(df_3)
dim(df_4)
dim(df_5)

# The value 119 is actually determined by the dimensions, just subtract 2 from what you get in dimensions

col_times <- paste(rep(1:182), sep = "") 

# Set column names for the dataframes 

colnames(df_1) <- c("file", "channel", col_times)
colnames(df_2) <- c("file", "channel", col_times)
colnames(df_3) <- c("file", "channel", col_times)
colnames(df_4) <- c("file", "channel", col_times)
colnames(df_5) <- c("file", "channel", col_times)

```

---

## Create annotation columns and add timepoints

Timepoints are treated as positions along line scan

```{r annotate}

# Helper function: pivot longer and annotate 
annotate_df <- function(df) {
  df[2:nrow(df), 1:184] %>% # 121 comes from dimensions
    separate_wider_delim(file, delim = "_",
                         names = c("date", "strain", "mRNA1", "mRNA2", "embryoID", NA)) %>%
    pivot_longer(cols = `1`:`182`, names_to = "xpoint", values_to = "intensity") %>%  #change the dimension value according to line 63
    mutate(unique_id = paste(date, strain, mRNA1, mRNA2, embryoID, channel, sep = "_"))
}

# Helper function: add timepoints from the header row of any loaded table
addInTimepoints <- function(tibble, ref_table) {
  timepoints <- as.vector(unlist(ref_table[1, 3:184])) #need the full original dimension
  repnumb    <- nrow(tibble) / 182 
  tibble::add_column(tibble, timepoints = rep(timepoints, repnumb), .after = "xpoint")
}

# Add timepoints as proxy for pixel position along line scan 
df_1_pt <- addInTimepoints(annotate_df(df_1), df_1) # Use first dataframe as reference 
df_2_pt <- addInTimepoints(annotate_df(df_2), df_1)
df_3_pt <- addInTimepoints(annotate_df(df_3), df_1)
df_4_pt <- addInTimepoints(annotate_df(df_4), df_1)
df_5_pt <- addInTimepoints(annotate_df(df_5), df_1)

# Sanity check 
head(df_1_pt)

```

---

## Merge all data and apply fixed coordinate alignment

L4440 controls are included once (shared reference across all experiments).
Peak alignment is skipped: every embryo receives the same fixed coordinate,
recentered so xpoint 1:333 maps to aligned_row -166:166.

```{r merge_align}

# Concatenate the dataframes 
all_data <- rbind(
  df_1_pt, df_2_pt, df_3_pt, df_4_pt, df_5_pt
)

center <- ceiling(182/2) # Using dims, find center point 
center

dt <- as.data.table(all_data)
dt$xpoint <- as.integer(dt$xpoint)
dt[, aligned_row := xpoint - center]

fwrite(dt, "/Users/ambikabasu/Nishimura_lab/Macros/04_R_analysis_tnpo/03_output/aligned_data.csv")

total_align_long <- dt %>%
  select(strain, aligned_row, unique_id, intensity)

cat("Total rows:", nrow(total_align_long), "\n")
cat("Unique embryo IDs:", n_distinct(total_align_long$unique_id), "\n")

head(total_align_long)

```

---

## Normalize

Each embryo is normalized to its own mean intensity (per embryo per channel).

```{r normalize}

norm_total <- total_align_long %>%
  separate_wider_delim(unique_id, delim = "_",
                       names = c("date", NA, "mRNA1", "mRNA2", "embryoID", "channel")) %>%
  group_by(date, embryoID, channel, mRNA1, mRNA2) %>%
  mutate(normalized_intensity = intensity / mean(intensity, na.rm = TRUE)) %>%
  ungroup()

treatment_order <- c("ch1", "ch2")
norm_total$channel <- factor(norm_total$channel, levels = treatment_order)

# ch1 = gapdh, ch2 = tnpo-1
norm_total <- norm_total %>%
  mutate(treatment_type = case_when(
    channel == "ch1" ~ "gapdh",
    channel == "ch2" ~ "tnpo-1",
    TRUE ~ NA_character_
  ))

table(norm_total$channel, norm_total$treatment_type)

```

---

## Color palettes

```{r palettes}

treatment_colors <- c(
  "ch2" = "#2166AC",   # tnpo-1 — dark blue
  "ch1" = "#4D4D4D"    # gapdh — dark gray
)
type_colors <- treatment_colors

channel_labels <- c("ch1" = "gapdh mRNA", "ch2" = "tnpo-1 mRNA")

```

---

## Section 1: Individual normalized line scans

All embryo-level line scans, faceted by channel and treatment.

```{r linescans_individual, fig.height = 8, fig.width = 10}

p_linescan_individual <- ggplot(norm_total,
       aes(x = aligned_row, y = normalized_intensity,
           group = interaction(date, embryoID),
           color = channel)) +
  geom_line(alpha = 0.6, linewidth = 0.3) +
  scale_color_manual(values = treatment_colors) +
  facet_wrap(~ channel, labeller = labeller(channel = channel_labels)) +
  labs(
    title = "Individual normalized line scans — all treatments",
    x     = "Position along line scan",
    y     = "Normalized intensity",
    color = "Treatment"
  ) +
  guides(color = "none") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
p_linescan_individual


# Save the plotted data
write.csv(
  norm_total %>%
    select(date, embryoID, channel, aligned_row, normalized_intensity),
  "individual_linescan_values.csv",
  row.names = FALSE
)

```

---

## Section 2: Mean +- SD line scans

### 2a. One panel per channel

```{r linescans_mean_overlaid, fig.height = 6, fig.width = 12}

mean_signal <- norm_total %>%
  group_by(aligned_row, channel) %>%
  summarize(
    mean_signal = mean(normalized_intensity, na.rm = TRUE),
    sd_signal   = sd(normalized_intensity,   na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  mutate(
    ymeanmax = mean_signal + sd_signal,
    ymeanmin = mean_signal - sd_signal
  )

ggplot(mean_signal,
       aes(x = aligned_row, y = mean_signal,
           group = channel)) +
  geom_ribbon(aes(ymin = ymeanmin, ymax = ymeanmax,
                  fill = channel), alpha = 0.2, color = NA) +
  geom_line(aes(color = channel), linewidth = 0.8) +
  scale_color_manual(values = treatment_colors) +
  scale_fill_manual(values  = treatment_colors) +
  facet_wrap(~ channel, labeller = labeller(channel = channel_labels)) +
  labs(
    title = "Mean +- SD line scans, per channel",
    x     = "Position",
    y     = "Mean normalized intensity"
  ) +
  guides(color = "none", fill = "none") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

```

---

## Section 3: log2 fold change enrichment

Ratio of normalized intensity at position 1 versus the mean of positions
-100 and +100. Higher values indicate more enriched localization signal
(consistent with membrane localization).

```{r}

peaks_and_valleys <- norm_total %>%
  filter(aligned_row %in% c(-50, 1, 50))

nest_pandv <- peaks_and_valleys %>%
  group_by(date, mRNA1, mRNA2, embryoID, channel) %>%
  nest()

my_calc2 <- function(df) {
  df$normalized_intensity[df$aligned_row == 1] /
    mean(df$normalized_intensity[df$aligned_row %in% c(-50, 50)])
}

foldChange_calc <- nest_pandv %>%
  mutate(fc_enrich = map_dbl(data, my_calc2)) %>%
  mutate(
    log2_fc        = log(fc_enrich, base = 2),
    treatment_type = case_when(
      channel == "ch1" ~ "gapdh",
      channel == "ch2" ~ "tnpo-1",
      TRUE ~ NA_character_
    ),
    channel        = factor(channel, levels = treatment_order)
  )

summary(foldChange_calc$log2_fc)

summary(foldChange_calc$log2_fc)

```

### 3a. Merged boxplot, x-axis by treatment, faceted by channel

```{r boxplot_merged, fig.height = 6, fig.width = 8}

foldChange_calc <- foldChange_calc %>%
  mutate(treatment_type = factor(treatment_type, levels = c("gapdh", "tnpo-1")))

imb2_ch2_ref <- foldChange_calc %>%
  ungroup() %>%
  filter(channel == "ch2") %>%
  summarize(med = median(log2_fc, na.rm = TRUE)) %>%
  mutate(channel = "ch2")

p_merged <- ggplot(foldChange_calc,
                   aes(x = treatment_type,
                       y = log2_fc,
                       fill = treatment_type)) +
  geom_boxplot(outlier.shape = NA, width = 0.6) +
  geom_jitter(aes(color = treatment_type),
              position = position_jitterdodge(jitter.width = 0.25),
              size = 1.5, alpha = 0.7) +
  scale_fill_manual(
    values = type_colors,
    guide  = "none"
  ) +
  scale_color_manual(
    values = c("imb-2" = "black", "set-3" = "black"),
    guide  = "none"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_y_continuous(limits = c(-0.5, 1.5)) +
  labs(
    title = "log2 enrichment at position 0 vs ±50",
    x     = "mRNA",
    y     = "log2 (intensity at 0 / mean intensity at ±50)",
    fill  = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 35, hjust = 1, face = "italic"),
    strip.text      = element_text(face = "bold"),
    legend.position = "none"
  )

p_merged

```

---

## Summary statistics

```{r summary_stats}

summary_stats_table <- foldChange_calc %>%
  group_by(treatment_type, channel) %>%
  summarise(
    n             = n(),
    min_log2fc    = round(min(log2_fc,    na.rm = TRUE), 3),
    median_log2fc = round(median(log2_fc, na.rm = TRUE), 3),
    mean_log2fc   = round(mean(log2_fc,   na.rm = TRUE), 3),
    max_log2fc    = round(max(log2_fc,    na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(channel, treatment_type)

print(summary_stats_table, n = Inf)

```

---

## Statistics - Pairwise Wilcoxon vs L4440

```{r pairwise_wilcox}

# Wilcoxon test comparing set-3 vs imb-2 enrichment
pairwise_stats <- foldChange_calc %>%
  ungroup() %>%
  wilcox_test(log2_fc ~ treatment_type) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

pairwise_stats

```

---

## Session info

```{r sessionInfo}
sessionInfo()
```

---

## Export plots and stats

```{r export_outputs}

today      <- format(Sys.Date(), "%y%m%d")
output_dir <- "../03_output"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(file.path(output_dir, paste0(today, "_boxplot_merged.svg")), plot = p_merged,
       width = 12, height = 7)
ggsave(file.path(output_dir, paste0(today, "_linescan_individual.svg")), plot = p_linescan_individual,
       width = 14, height = 12)
ggsave(file.path(output_dir, paste0(today, "_linescan_mean.svg")), plot = p_linescan_mean,
       width = 14, height = 12)

write_csv(summary_stats_table, file.path(output_dir, paste0(today, "_summary_stats.csv")))
write_csv(pairwise_stats, file.path(output_dir, paste0(today, "_pairwise_wilcoxon_stats.csv")))

cat("Exported to:", normalizePath(output_dir), "\n")

```
